Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies

### 2012-07-27 MIMOSA.tex
Update the manuscript and supp document.

\title[MIMOSA: Combinatorial Mixture Models or Single--Cell Assays]{Combinatorial Mixture Models for Single--Cell Assays with Application to Vaccine Studies}
%\author[1]{Greg Finak}
%\author[2]{SC De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}

@@

\pagerange{\pageref{firstpage}--\pageref{lastpage}} 
\volume{ }
\pubyear{ }
\artmonth{ }

@@

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single--cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays. Our model allows the inference specific to each observation, while borrowing strength across observations through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods. Using a simulation study we show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model. 
\end{abstract}

\begin{keywords}
Mixture modelling, hierarchical modelling, bayesian modelling, single--cell assays, immunology
\end{keywords}

@@

\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. 
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. 
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. 
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single--cell measurement of multiple surface and intracellular markers (up to 17 markers). 
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. 
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. 
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}. 
This heterogeneity may carry important information, thus single--cell expression data should not be analyzed in the same fashion as cell-population level data. 
Special treatment of single--cell level data, which preserves information about population heterogeneity, is warranted in general. 
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. 
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. 
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single--cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
Using this matrix, one can form $2^K$ putative cell subsets. 
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).

@@

\textit{Flow cytometry}:
The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see Web Appendix A). 
The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 individual from the vaccinee group at two time points, day 0 and day 182. Samples on day 0 were taken just before vaccination and no response is expected there. 
The corresponding samples can be used as negative controls. 
Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. 
The data is described in Web Appendix A.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen individuals. T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post--stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls.~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al. could you fill this in?}.

@@

For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Individuals that show differential expression for a given marker will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.

@@

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder. In some study, such as the ICS studies used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix B and we refer to it as the one-sided model.

@@

\subsection{Parameter estimation}
\label{s:estimation}
Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed forms (Web Appendix C), and are given by,
 \begin{align*}
 \begin{split}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
	=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
	\end{split}
 \end{align*}

and
\begin{align}
	\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
\label{model2:unconstrained}
\end{split}
\end{align}
Assuming that the missing data, the $z_i$'s, are known, we define the complete data log-likelihood as follows,
\begin{equation}
\begin{split}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w)\label{eq:cll}
\end{split}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated. In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed form but can easily be estimated. All calculations for the one-sided case are described in Web Appendix B.

@@

\subsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Web Appendix D. We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold--change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.

%The constrained model was applied to an ICS data set from a real--world vaccine trial in order to identify responders to antigen stimulation. The unconstrained model was applied to Fluidigm single--cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single--cells. We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold--change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+? Didn't look at CD8's}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold--change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave better estimates of the observed false discovery rate than competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
    \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=.6\columnwidth]{Figures/2}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
\end{tikzpicture}
 };
 \node [below=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=.6\columnwidth]{Figures/1}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
 };
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one--sided model) and competing methods on ICS data from HVTN065. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T--cells stimulated with A--B) ENV-1-PTEG and expressing IFNg or C--D) ENV-1-PTEG and expressing IL2. ROC and FDR plots of other cytokine combinations can be found in Web Figure A.}
\label{fig:HVTN065}
\end{figure}

\subsection{Single-cell gene expression}
We applied the MIMOSA model to Fluidigm single-cell gene expression. This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation. In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm} A,B). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). A similar analysis using a two-sided Fisher's exact test and clustering the signed q--values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation--specific patterns. Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.

\begin{figure}
\centering
\begin{tikzpicture} [auto,node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
\node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of B] (C) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
\node at (0,1) [font=\tiny\sffamily] {C} ;
\end{scope}
\end{tikzpicture}
};
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single--cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The posterior differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Clustering of the signed q--values from Fisher's exact test. Genes selected from Fisher's exact test at the 10\% FDR level. }
\label{fig:fluidigm}
\end{figure}

@@

%We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles. As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations. We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T--cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}). 

>>>>

\title[MIMOSA: Combinatorial Mixture Models or Single-Cell Assays]{Combinatorial Mixture Models for Single-Cell Assays with Application to Vaccine Studies}
%\author[1]{Greg Finak}
%\author[2]{SC De Rosa}
%\author[3]{Mario Roederer}
%\author[1]{Raphael Gottardo}

@@

\pagerange{\pageref{firstpage}-\pageref{lastpage}} 
\volume{ }
\pubyear{ }
\artmonth{ }

@@

\begin{abstract}
Most biological samples (\textit{e.g.}, blood and tissue) are composed of many functionally distinct cell subsets that can only be measured accurately using single-cell assays. The characterization of these small cell subsets is crucial to decipher system level biological changes. For this reason, an increasing number of studies rely on single-cell assays to provide single-cell measurements of multiple genes and proteins from bulk cell samples. A common problem in the analysis of such data is to identify markers (or combinations of markers) that are differentially expressed between two biological conditions (\textit{e.g.}, before/after vaccination), where expression is defined as the proportion of cells expressing that marker (or marker combination) in the cell subset(s) of interest.
Here we present a Bayesian hierarchical framework based on beta-binomial mixture models for testing for differential marker expression using such single-cell assays. Our model allows the inference specific to each observation, while borrowing strength across observations through common prior distributions. We propose two approaches for parameter estimation: an empirical-Bayes approach using an Expectation-Maximization algorithm and a fully Bayesian one based on a Markov chain Monte Carlo algorithm. We compare our method against Fisher's exact test, a likelihood ratio test, and basic log-fold changes. Using several experimental immunological assays measuring proteins or genes at the single-cell level, we find that our method has higher sensitivity and specificity than alternative methods. Using a simulation study we show that our framework is also robust to model misspecification. Finally, we also demonstrate how our approach can be extended to testing multivariate differential expression across multiple marker combinations using a multinomial-Dirichlet model. 
\end{abstract}

\begin{keywords}
Mixture modelling, hierarchical modelling, bayesian modelling, single-cell assays, immunology
\end{keywords}

@@

\section{Introduction}
\label{s:intro}
Cell populations, particularly in the immune system, are never truly homogeneous; individual cells may be in different biochemical states that define functional but measurable differences between them. 
This single-cell heterogeneity is informative, but lost in assays that measure cell mixtures. 
For this reason, endpoints in vaccine and immunological studies are measured through a variety of assays that provide single-cell measurements of multiple genes and proteins. 
In the 1970s, single-cell analysis was revolutionized with the development of fluorescence-based flow cytometry (FCM). 
Since then, instrumentation and reagent advances have enabled the study of numerous cellular processes via the simultaneous single-cell measurement of multiple surface and intracellular markers (up to 17 markers). 
More recent technological development have drastically extended the capabilities of single-cell cytometry to measure dozens of simultaneous parameters per cell~\citep{Bendall:2011wf}. 
Although cells sorted using well-established surface markers may appear homogeneous, mRNA expression of other genes within these cells can be heterogeneous~\citep{Narsinh:2011gn, Flatz:2011jb} and could further characterize and subset these cells. 
A new technology based on microfluidic arrays combined with multiplexed polymerase chain reactions (PCR) can now be used to perform thousands of PCRs in a single device, enabling simultaneous, high-throughput gene expression measurements at the single-cell level across hundreds of cells and genes~\citep{Pieprzyk:2009uc}. 
While classic gene expression microarrays sum the expression from many individual cells, the intrinsic stochastic nature of biochemical processes results in relatively large cell-to-cell gene expression variability~\citep{vanOudenaarden200915a}. 
This heterogeneity may carry important information, thus single-cell expression data should not be analyzed in the same fashion as cell-population level data. 
Special treatment of single-cell level data, which preserves information about population heterogeneity, is warranted in general. 
For this reason, single-cell assays are an important tool in immunology, providing a functional and phenotypic snapshot of the immune system at a given time. 
These assays typically measure multiple markers simultaneously on individual cells in a heterogeneous mixture such as whole blood or peripheral blood cell mononuclease (PBMC), and are used for immune monitoring of disease, vaccine research, and diagnosis of haematological malignancies~\citep{Altman:1996wf,Betts:2006dw,Inokuma:2007tn}.

During analysis, cell level marker intensities are typically thresholded as positive or negative so that subsets with different multivariate $+/-$ combinations can be obtained as Boolean combinations. 
For some assays (\textit{e.g.}, flow cytometry), the positivity thresholds are set based on prior biological knowledge while for others thresholds are given by the assay technology. 
This is the case for the Fluidigm technology where genes are recorded as absent (not expressed) or present (expressed) at the single-cell level.
 After this thresholding step, we obtain a Boolean matrix of dimension $N\times K$ where $N$ is the number of cells recorded and $K$ is number of markers. 
Using this matrix, one can form $2^K$ putative cell subsets. 
When $K$ is large there is a combinatorial explosion of the number of subsets, and many of these might be small or even empty. 
A common statistical problem is, for a given marker combination, to identify individuals for whom the proportion of cells expressing that combination is significantly different between two experimental conditions (\textit{e.g.}, before and after vaccination).

@@

\textit{Flow cytometry}:
The primary ICS data set is from a  trial testing the GeoVax DNA and MVA vaccines in a prime-boost regimen with 120 individuals (98 vaccines and 20 placebo recipients, see Web Appendix A). 
The goal of this data set was to assess the immune response to the vaccine across multiple stimulations, time points, cytokines and T-cell subsets. 
Here, we analyze a subset of the data consisting of 98 individual from the vaccine group at two time points, day 0 and day 182. For ease of presentation we restricted ourselves to the CD4+ T-cell subsets. Samples on day 0 were taken just before vaccination and no response is expected there. The corresponding samples can be used as negative controls. 
Conversely, day 182 (26 weeks) should be close to the immunogenicity peak, and many individuals are expected to respond, for some cytokines at least. More details can be found in Web Appendix A.

\textit{Fluidigm single-cell gene expression}: This is a single-cell gene expression data set of sorted CD8$+$ T-cells from sixteen individuals. T-cells isolated by flow cytometry from sixteen individuals were stimulated in blocks of four individuals with four different antigens (HIV Gag, HIV Nef, CMV pp65 tm10, CMV pp65 nlv5) and gene expression post-stimulation measured at the single-cell level using the BioMark system (Fluidigm) $96 \times 96$ well arrays. The expression from the simulated samples  was compared to paired, unstimulated controls~\footnote{Add a bit more info here, how many genes, tetramer sorted, etc, Mario et al. could you fill this in?}.

@@

For a given individual and stimulation, we consider a marker to be differentially expressed if the proportion of positive cells in the stimulated samples is different from the number of positive cells in the un-stimulated sample. Individuals that show differential expression will be called responders for that marker. In this section, we shall be concerned with identifying differential expression one marker at a time, using a beta-binomial mixture model as described in what follows.

@@

\begin{equation*}
(n_{si}|p_{si}) \sim \mathrm{Bin}(N_{si},p_{si})\quad \text{and}\quad (n_{ui}|p_{ui}) \sim \mathrm{Bin}(N_{ui},p_{ui})\label{eq:bino_likelihood}
\end{equation*}
where $p_{si}$, $p_{ui}$ are the unknown proportions for the stimulated and un-stimulated paired samples. In order to detect responding individuals we consider two competing models:
\begin{equation*}
{\cal M}_0: p_{ui}=p_{si}\quad \text{and}\quad {\cal M}_1: p_{ui}\ne p_{si}. \label{eq:models}
\end{equation*}
Under the null model, ${\cal M}_0$, there is no difference between the stimulated and un-stimulated samples, and the proportions are equal. Under the alternative model, ${\cal M}_1$, there is a difference in proportions between the two samples and the individual $i$ is a responder. In some study, such as the ICS data used here, the proportion of positive cells is expected to only increase after stimulation, in which case the alternative model should be defined as $p_s>p_u$. This alternative parametrization is described in Web Appendix C and we refer to it as the one-sided model.

@@

\subsection{Parameter estimation}
\label{s:estimation}
Our estimation algorithms make direct use of the marginal likelihoods, $\mathrm{L}_0$ and $\mathrm{L}_1$, obtained after integrating out the $p_{\{s,u\}i}$'s for the null and alternative models, which greatly simplify our calculations. Given the conjugacy of the priors, the marginal likelihoods $\mathrm{L}_0$ and $\mathrm{L}_1$ are available in closed-forms (Web Appendix B), and are given by,
 \begin{align*}
 \begin{split}
  	\mathrm{L}_0(\alpha_u,\beta_u|\mathbf{y})
	=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}}\binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{si}+n_{ui}+\alpha_u,N_{si}-n_{si}+N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}
	\end{split}
 \end{align*}

and
\begin{align}
	\begin{split}
\mathrm{L}_1(\alpha_u,\beta_u,\alpha_s,\beta_s|\mathbf{y}) 
=&\prod_{i=1}^P\binom{N_{ui}}{n_{ui}} \binom{N_{si}}{n_{si}}\cdot\\ &\frac{\mathrm{B}(n_{ui}+\alpha_u,N_{ui}-n_{ui}+\beta_u)}{\mathrm{B}(\alpha_u,\beta_u)}\cdot \\ &\frac{\mathrm{B}(n_{si}+\alpha_s,N_{si}-n_{si}+\beta_s)}{\mathrm{B}(\alpha_s,\beta_s)}
\label{model2:unconstrained}
\end{split}
\end{align}
Assuming that the missing data, the $z_i$'s, are known, we define the complete data log-likelihood as follows,
\begin{equation}
\begin{split}
l(\boldsymbol{\theta}|\mathbf{y},\mathbf{z})=\sum_i z_i l_0(\alpha_u, \beta_u|\mathbf{y}_i) +(1-z_i) l_1(\alpha_u, \beta_u, \alpha_s, \beta_s|\mathbf{y}_i)+\\z_i\log(w)+(1-z_i)\log(1-w)\label{eq:cll}
\end{split}
\end{equation}
where $l_0$ and $l_1$ are the log marginal-likelihoods and $\boldsymbol{\theta}\equiv(\alpha_u, \beta_u, \alpha_s,\beta_s, w)$ is the vector of parameters to be estimated. In the one-sided case, the alternative prior specification must satisfy the constraint $p_s>p_u$, and the marginal likelihood derivation involves the calculation of a normalizing constant that is not available in closed-form but can easily be estimated. All calculations for the one-sided case are described in Web Appendix C.

@@

\subsection{MCMC algorithm}
Realizations were generated from the posterior distribution via MCMC algorithms~\citep{Gelfand:1996wc}. All updates were done via Metropolis-Hastings sampling except for the $z_i$'s and $w$ that were done via Gibbs samplings.
Details about the algorithms are given in Web Appendix B. We used the method of \cite{Raftery:1992vp,Raftery:1996ws} to determine the number of iterations, based on a short pilot run of the sampler. For each dataset presented here, this suggested that a sample of no more than about 1,000,000 iterations with 50,000 burn-in iterations was sufficient to estimate standard posterior quantities. Guided by this, and leaving some margin, we used 2,000,000 iterations after 50,000  burn-ins for each dataset explored here.

\section{Results}
\label{s:results}
In this section, we apply our MIMOSA model to the data described in Section \ref{s:data}, and present the results of a simulation study based on the ICS data. The performance of MIMOSA was evaluated and compared against Fisher's exact test, the likelihood ratio test, and log fold-change by ROC (receiver operator characteristic) curve analysis and by comparing the observed FDR (false discovery rate) against the nominal FDR for each data set.

%The constrained model was applied to an ICS data set from a real-world vaccine trial in order to identify responders to antigen stimulation. The unconstrained model was applied to Fluidigm single-cell gene expression data to identify genes differentially expressed between stimulated and unstimulated conditions in populations of single-cells. We also performed simulation studies to assess the performance of the constrained and unconstrained models in a univariate and multivariate settings.

\subsection{ICS}
Using the ICS data, an ROC (receiver operator characteristic) analysis was performed to assess the sensitivity and specificity of the one-sided MIMOSA model compared to a one-sided Fisher's exact test, log fold-change, and a likelihood ratio test based on the MIMOSA model for identifying vaccine responders and non-responders. Observations at the day 0 time point were treated as true negatives, while observations at the day 182 time point were treated as true positives (potentially underestimating the sensitivity due to non-responders). We examined the CD4+ T-cell cytokine responses \footnote{What about CD8+? Didn't look at CD8's}. The MIMOSA model has higher sensitivity and specificity than Fisher's exact test, the likelihood ratio test, or log fold-change for discriminating vaccine responders and non-responders  as shown by the ROC curves on Figure~\ref{fig:HVTN065} A. 
In addition, MIMOSA gave estimates of the observed false discovery rate that are better or comparable to competing methods (Figure~\ref{fig:HVTN065} B). These results are consistent for other cytokines (see Web Figure A).

\begin{figure} %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture} [auto, node distance=0cm]
\node at (0,0) (A){
    \begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) (foo) {\includegraphics[width=.6\columnwidth]{Figures/2}};
    \begin{scope}[x={(foo.south east)},y={(foo.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {A} ;
    \node at (0.5,1) [font=\tiny\sffamily] {B} ;
    \end{scope}
\end{tikzpicture}
 };
 \node [below=of A] (B) {
 \begin{tikzpicture}
    \node[anchor=south west, inner sep=0] at (0,0) (bar){\includegraphics[width=.6\columnwidth]{Figures/1}};
    \begin{scope} [x={(bar.south east)},y={(bar.north west)}]
    \node at (0,1) [font=\tiny\sffamily] {C} ;
    \node at (0.5,1) [font=\tiny\sffamily] {D} ;
    \end{scope}
    \end{tikzpicture}
 };
\end{tikzpicture}
   \caption{Performance of MIMOSA (EM and MCMC implementations, one-sided model) and competing methods on ICS data from HVTN065. Sensitivity and specificity (ROC analysis) as well as observed and nominal false discovery rates for positivity calls from CD4+ T-cells stimulated with A-B) ENV-1-PTEG and expressing IFNg or C-D) ENV-1-PTEG and expressing IL2. ROC and FDR plots of other cytokine combinations can be found in Web Figure A.}
\label{fig:HVTN065}
\end{figure}

\subsection{Single-cell gene expression}
We applied the MIMOSA model to Fluidigm single-cell gene expression. This time we used the two-sided MIMOSA model because genes could be up or down regulated upon stimulation. In order to detect stimulation specific changes of expression, our model was fit to each gene within each stimulation. The results presented in Figure~\ref{fig:fluidigm} show that MIMOSA identifies stimulation-specific differences in the proportions of cells expressing each gene, while preserving inter-individual variability (Figure~\ref{fig:fluidigm} A,B). These patterns are evident in the  posterior probabilities (Figure~\ref{fig:fluidigm} A), and preserved in the posterior estimates of the differences of proportions (Figure~\ref{fig:fluidigm} B). A similar analysis using a two-sided Fisher's exact test and clustering the signed q-values (Figure~\ref{fig:fluidigm} C) does not reveal any stimulation-specific patterns. Comparing the size of the gene sets, Fisher's exact test identified 47 significant genes while MIMOSA identified 50 significant genes, with 39 genes identified in common between the two methods.

\begin{figure}
\centering
\begin{tikzpicture} [auto,node distance=0cm]
\node at (0,0) (A) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (foo) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot1.pdf}};
\begin{scope} [x={(foo.south east)},y={(foo.north west)}]
\node at (0,1) [font=\tiny\sffamily] {A} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of A] (B) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (bar) {\includegraphics[width=0.29\columnwidth]{Figures/Fluidigm_plot2.pdf}};
\begin{scope} [x={(bar.south east)},y={(bar.north west)}]
\node at (-0.1,1.1) [font=\tiny\sffamily] {B} ;
\end{scope}
\end{tikzpicture}
};
\node [right=of B] (C) {
\begin{tikzpicture}
\node[anchor=south west, inner sep=0] at (0,0) (baz) {\includegraphics[width=0.29\columnwidth]{Figures/FluidigmFisher.pdf}};
\begin{scope} [x={(baz.south east)},y={(baz.north west)}]
\node at (0,1) [font=\tiny\sffamily] {C} ;
\end{scope}
\end{tikzpicture}
};
\end{tikzpicture}
\caption{Signed posterior probability, difference and log-odds ratio of the proportion of single-cells expressing each gene on a 96x96 Fluidigm array. The posterior probability of response times the sign of the change in expression is shown in A) (red indicates a decrease, green an increase, relative to the control). Columns and rows are clustered based on these signed posterior probabilities. B) The posterior differences in proportion of cells expressing a gene in the stimulated vs. control samples. Rows and columns are ordered as in A) for comparison. The traces show the deviations of each cell from zero. Colors along the columns denote different stimulations (green: CMV pp65 nlv5, red: HIV Gag, orange: HIV Nef, yellow: CMV pp65 tm10). C) Clustering of the signed q-values from Fisher's exact test. Genes selected from Fisher's exact test at the 10\% FDR level. }
\label{fig:fluidigm}
\end{figure}

@@

%We applied the Multinomial-Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles. As a proof of principle we limited the model to the two-component case (Figure~\ref{tab:nesting}), with one component for the null model where we expect no difference between stimulations, and one component for the alternative model where we expect a difference between stimulations. We fit the model to the IL4 and TNFa cytokines for Env3 and Env1 stimulated CD4 T-cells from the day zero and day 28 time points (Figures~\ref{fig:polyfunctionality} and \ref{fig:polyfunctionalityenv1}). 
